A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-05-19 papers/contact/paper.tex
clean up description of contact functionals Tested everything on bingley.

For the symmetric $S$ case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_0(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \\
  n_0(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_0$ is one of the fundamental measures
of Fundamental Measure Theory (FMT), and seems ideal for treating
touching spheres.

And for the asymmetric $A$ case:
\begin{align}
  g^A_\sigma(\rr) &= \frac{1}{n(\rr)n_A(\rr)}
  \int n^{(2)}(\rr, \rr + \rr')
  \frac{\delta(|\rr'| - \sigma)}{4\pi\sigma^2}d\rr' \\
  n_A(\rr) &= \int n(\rr')
  \frac{\delta(|\rr-\rr'|-\sigma)}{4\pi\sigma^2} d\rr'
\end{align}
In this case, the two-particle density is averaged over spheres
touching a sphere that is aat the given position $\rr$.

@@

In order to motivate the derivation of our functional for contact
density, we begin by deriving the well-known contact density for a
homogeneous hard-sphere fluid from the Carnahan-Starling free energy.
The contact density is most easily computed from the contact density
theorem, which states that the pressure on any hard surface is
determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact}.
\end{align}
Since we are interested in the contact density at the surface of the
hard spheres, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the Carnahan-Starling free energy on
hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{3} R^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi \sigma^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi \sigma^2} \frac{dA_{HS}}{dR} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:p_{HS}^A}\\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  g_\sigma^A(\rr) &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
Equation~\ref{eq:g-A-exact} is an exact expression for $g_\sigma^A$
given the exact functional $A_{HS}$.  In practice, we use the
approximate White Bear functional, and find an approximate expression
for $g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve more convolutions than are present in the free
energy itself, making this formulation relatively expensive.

@@

\section{Gross's asymmetrically averaged correlation functional}\label{sec:gross}
We compare with the averaged correlation function of
Gross\cite{gross2009density}, which is of the asymmetrical variety
($g_\sigma^A$), where the correlation function is averaged at location
of one of the spheres in concact.
\begin{align}
  g_\sigma^\text{Gross,A} &= \frac{1 - \frac12 n_{3'}}{\left(1 -
    n_{3'}\right)^3} \\
  n_{3'}(\rr) &= \int n(\rr')\Theta(|\rr-\rr'| - 2R) d\rr'
\end{align}
where $n_{3'}$ is an averaged
filling fraction defined in a manner analogous to the $n_3$ of FMT,
but with the hard-sphere radius replaced by the hard-sphere diameter,
so the integrals range over spheres that are twice as large.

@@

An alternate definition of the averaged correlation function for an
inhomogeneous system would be the correlation function averaged for
spheres \emph{touching at a given point}.  This would be a symmetric
measure, since both spheres involved are treated in the same way.  In
fact, the density of such spheres is just the density $n_0(\rr)$ from
FMT.  See Figure~\ref{fig:n0} for an illustration $n_0$.  Since $n_0$
is the fundamental measure of sphere surface contact, it makes sense
to define a corresponding correlation function $g_\sigma^S$.  This is
\emph{almost} what is done by Yu and  Wu

\fixme{Analogous to the assymetric case is the symmetric, for which 
we are no longer able to simply pick a hypothetical sphere centered 
at position $r$ and expand its radius.  Instead we must formulate a way 
to expand the radii of spheres that just touch a specific point of 
contact and only expand them at that point of contact, in a symmetric 
fashion.  Therefore, when we compute the neccasary derivitives wih 
respect to $R$', we expand a multitude of spheres, all of them positioned 
around and just touching the point $\mathbf{r}$.  When their radii 
expand, each does not expand outward in all directions as in the 
case of A, rather they each expand outward at only the point $\mathbf{r}$.  One can imagine a shell that touches $\mathbf{r}$ that has an infinitesimal bit of shell area protruding outward at $\mathbf{r}$ while the rest of its shell remains 
at radius $R$.  When we expand each sphere that touches $\mathbf{r}$ 
in this peculiar fashion, we have in the end the same amount of shell 
area protruding outward as we would if we had simply expanded one 
sphere in fashion of case A.  Also the protrusions, once added up 
for all spheres sourrounding $\mathbf{r}$, will yield a symmetry 
in their direction of protrusion.  Each protruding sphere bit will 
expand outward from its center and all directions will be covered uniformly.
See Figure \ref{fig:contact}.

\begin{figure}
\includegraphics[width=5cm]{figs/contact}
\caption{Set of hard spheres that included in $n_0(\mathbf{r})$, which
  consist of those which just touch the point $\mathbf{r}$.}
\label{fig:contact}
\end{figure}

These qualitative concepts lead us to derivations for the $\frac{\delta A_{HS}}{\delta R(\mathbf{r})}$ term in our derivation of the pressure, eq \ref{eq:p_{HS}}.  The rest of the derivation results when we recognize that instead of dividing the pressure by $n(\mathbf{r})$ and $n_A(\mathbf{r})$ as in the assymmetric case, we will instead divide by $n_0(\mathbf{r})^2$ since we have symmetry around the point $\mathbf{r}$.  This will yield our symmetric correlation, $g_\sigma^S$:

\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:p_{HS}^S}\\
  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  g_\sigma^A(\rr) &= \frac{1}{n_0(\rr) n_0(\rr)}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}}Complete expressions for $g_\sigma^S$ are included in
Appendix~\ref{appendix:g-S}.\\

\section{Yu and Wu's functional}\label{sec:yuwu}

Yu and Wu developed a functional that should correspond to the contact
density at a given contact
position~$g^a$~\cite{yu2002fmt-dft-inhomogeneous-associating}.  We
will compare both their functional as well as ours with simulation
data.  Yu and Wu use a symmetric formulation for $g_\sigma$, but
instead of using $n_0$ as the associated density, it uses a density
given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$ is
given by:
\begin{align}
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac32 \frac{R n_2}{3}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{R n_2}{3}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
I think the adjusted functional $g_\sigma^{\text{Yu,}S}$ can be directly
compared with our $g_\sigma^S$.  It is implemented in our code.

>>>>

For the symmetric $S$ case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_0(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \label{eq:gS} \\
  n_0(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_0$ is one of the fundamental measures
of Fundamental Measure Theory (FMT), and seems ideal for treating
touching spheres, as illustrated in Figure~\ref{fig:n0}.

And for the asymmetric $A$ case:
\begin{align}
  g^A_\sigma(\rr) &= \frac{1}{n(\rr)n_A(\rr)}
  \int n^{(2)}(\rr, \rr + \rr')
  \frac{\delta(|\rr'| - \sigma)}{4\pi\sigma^2}d\rr' \label{eq:gA} \\
  n_A(\rr) &= \int n(\rr')
  \frac{\delta(|\rr-\rr'|-\sigma)}{4\pi\sigma^2} d\rr' \label{eq:nA}
\end{align}
In this case, the two-particle density is averaged over spheres
touching a sphere that is aat the given position $\rr$.

@@

In order to motivate the derivation of our functional for contact
density, we begin by deriving the well-known contact density for a
homogeneous hard-sphere fluid from the Carnahan-Starling free energy.
The contact density is most easily computed from the contact density
theorem, which states that the pressure on any hard surface is
determined by the density at contact:
\begin{align}
  p &= k_BT n_\textit{contact}.
\end{align}
Since we are interested in the contact density at the surface of the
hard spheres, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the Carnahan-Starling free energy on
hard sphere radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
where $\eta \equiv \frac{4\pi}{3} R^3 n$ is the filling fraction.  We
can thus readily compute the derivative of free energy with respect to
hard-sphere radius:
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  \derivation{
    &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
    &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
    \\
  }
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi \sigma^2$, since the relevant area
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi \sigma^2} \frac{dA_{HS}}{dR} \\
  \derivation{
    &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
    &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 -
      \frac{\eta}2}{(1-\eta)^3} \\
  }
  &= k_BT n \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
Using the contact-value theorem, we can thus find the contact density
and the well-known correlation function evaluated at contact.
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3} \label{eq:cs-g}
\end{align}
Thus we can readily derive the contact density for the homogeneous
hard-sphere fluid.  Extending this to the inhomogeneous fluid will
require that we use density-functional theory.

@@

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.  Since the hard-sphere radius enters the FMT free energy
by means of the fundamental measures, this requires that we adjust the
convolutions to have a radius $R$ which depends on the location of the
sphere:
\begin{align}
  n_3^A(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-R(\rr'))d\rr' \label{eq:n3A}
\end{align}
Once we have done this---and similarly for the other fundamental
measures---we derive a formula for the correlation function
$g_\sigma^A(\rr)$ by finding the pressure on the surface of any hard
spheres located at position $\rr$, and using the contact-density
theorem to extract the contact density, from which we can establish
the averaged correlation function at contact.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:p_{HS}^A}\\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
The functional derivative $\frac{\delta A_{HS}}{\delta R(\mathbf{r})}$
in the above formula must be computed with the asymmetric formulation
for the fundamental measures.  In practice, we use the approximate
White Bear functional to find an approximate expression for
$g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve convolutions of derivatives of the free
energy, making this formulation computationally more expensive than
the free energy itself.

@@

@@

We now address the symmetrically averaged correlation function, which
is defined in Equation~\ref{eq:gS}.  This corresponds to the
correlation function averaged for
spheres \emph{touching at a given point}.  It is a symmetric
measure, since both spheres involved are treated in the same way.  In
fact, the density of such spheres is just the density $n_0(\rr)$ from
FMT.  See Figure~\ref{fig:n0} for an illustration $n_0$.  Since $n_0$
is the fundamental measure of sphere surface contact, it makes sense
to define a corresponding correlation function $g_\sigma^S$.  This is
\emph{almost} what is done by Yu and  Wu in
reference\cite{yu2002fmt-dft-inhomogeneous-associating}, as will be
discussed in Section~\ref{sec:yuwu}.

The procedure for finding the symmetrically averaged correlation
function $g_\sigma^S(\rr)$ is less straight-forward than 
\begin{align}
  n_3^S(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-R(\rr))d\rr'
\end{align}
The hard-sphere radius $R$ in this weighted density has an unusual
position dependence:  it depends on the position at which we are
evaluating the weighted density, rather than the position of the hard
spheres themselves.  This has the effect of expanding spheres as they
touch the point $\rr$, a concept illustrated in
Figure~\ref{fig:contact}.  This enables a similar derivation to that
which gave us $g_\sigma^A(\rr)$:
\begin{figure}
\includegraphics[width=5cm]{figs/contact}
\caption{Set of hard spheres that included in $n_0(\mathbf{r})$, which
  consist of those which just touch the point $\mathbf{r}$.}
\label{fig:contact}
\end{figure}
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n_0(\rr) 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:p_{HS}^S}\\
  \ncontact(\rr) &= \frac{1}{n_0(\rr) k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_0(\rr)} \\
  &= \frac{1}{n_0(\rr)^2}\frac{1}{ k_BT 4\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta R(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
where the functional derivative $\frac{\delta A_{HS}}{\delta
  R(\mathbf{r})}$ in this case is computed with the symmetric
formulation of the fundamental measures.  Complete expressions for
$g_\sigma^S$ are included in Appendix~\ref{appendix:g-S}.\\

\section{Gross's asymmetrically averaged correlation functional}\label{sec:gross}
One existing approximation for the ccorrelation function is that of
Gross\cite{gross2009density}, which is of the asymmetrically averaged
variety ($g_\sigma^A$):
\begin{align}
  g_\sigma^\text{Gross,A}(\rr) &= \frac{1 - \frac{2\pi}{3}R^3n_A(\rr)}{\left(1 -
    \frac{4\pi}{3}R^3n_A(\rr)\right)^3}
\end{align}
where $n_A$ is the averaged density defined in Equation~\ref{eq:nA}.
This formula is arrived at by using the density averaged over all
spheres that could be touching a sphere at point $\rr$ in the
Carnahan-Starling equation for the correlation function at contact,
given in Equation~\ref{eq:cs-g}.

\section{Yu and Wu's symmetrically averaged functional}\label{sec:yuwu}

\fixme{I (DJR) haven't yet edited this section!}
Yu and Wu developed a functional that should correspond to the contact
density at a given contact
position~$g^a$~\cite{yu2002fmt-dft-inhomogeneous-associating}.  We
will compare both their functional as well as ours with simulation
data.  Yu and Wu use a symmetric formulation for $g_\sigma$, but
instead of using $n_0$ as the associated density, it uses a density
given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$ is
given by:
\begin{align}
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac32 \frac{R n_2}{3}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{R n_2}{3}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
I think the adjusted functional $g_\sigma^{\text{Yu,}S}$ can be directly
compared with our $g_\sigma^S$.  It is implemented in our code.
